---
title: "Weather hazards probability"
subtitle: "Anti-weather preparation reference"
author: "Michael Mikhailidi"
date: "3/10/2016"
output: html_document
---
## Summary
 Document contains exploratory analyze of National Weather Service Storm database. Study provides data in format suitable for decision makers and give on weather phenomena probability and consequences for people and property. As a summary of the summary: _tornado_ is the most harmful event for people, while _TODO there_ bring the highest property damage.

## Data preparation and enreachment
I have used NOAA Storm database data, which will be transformed for visualization and understanding convenience

### Data load
 This study is based on National Weather Service data. Archive with data set was provided as part of Reproducible Research Project and could be found by in form of [BZ2 archive][https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). Last time data set was accessed at _Sunday, March 26, 2016 12:00 PM EST_.

To perform data  analysis code below downloads , checks if CSV data available in work directory and loads source into _weather_ data frame. Code below performs all necessary steps to prepare original data.

```{r dataload, echo=TRUE, message=FALSE, cache=TRUE} 
#Libraries required 
require(ggplot2)
require(dplyr)


# Source data  is not available in work directory.
if (!file.exists("weather.csv.bz2")) {
  message("Downloading data.") 
  dmethod <- 
  download.file(url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile="weather.csv.bz2", quiet = TRUE,
                method = switch(Sys.info()["sysname"], Linux="curl", Windows="wininet"))
}
# Load Soruce data.
# As son as object is quite heavy make sure it's not loaded.
if ( ! (is.object("weather") && 
       (nrow(weather) == 902297 && length(weather == 37)))) { 
  message("Reading data.")   
  weather<-read.csv("weather.csv.bz2",stringsAsFactors = FALSE )
  
}
```

Variables are described by [code book](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) as well, as separate [event description](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf) document. Let's examine loaded data and check how tidy data structure is.

```{r}
str(weather)
```

### Data extract and preparation 
From source data we extract weather events with summary of total events by type and how harmful those events for population (both injuries and fatalities).

```{r population}
require(ggplot2)
require(dplyr)
require(knitr)
events<-group_by(weather,EVTYPE) %>% summarize(total.injures=sum(INJURIES+FATALITIES),occurs=n()) 
byharm<-events[head(order(events$total.injures,decreasing = TRUE),10),]
bycount<-events[head(order(events$occurs,decreasing = TRUE),10),]
```
## Threats for population
Diagram shows top 10 most danjerous weather  for people life and health. Color reflectsoccurances  weather occurs for  for observation period.
```{r popgraph, dpi=120,fig.width=7,fig.height=7}
fills <- factor(findInterval(byharm$occurs,quantile(byharm$occurs,c(0,.5,1))),labels=c("rare","occasional","often"))
names(fills)<-
g<-ggplot(data=byharm,aes(x = EVTYPE,y=total.injures,fill=fills))
g<-g+geom_bar(stat="identity",position = position_dodge())
g<-g+theme(legend.title=element_blank())+scale_fill_brewer(palette="Set1",direction = -1)
g<-g+ggtitle("People injured by weather event type")+xlab("Weather Events")+ylab("People injured")
g<-g+theme(plot.title = element_text(lineheight=.9, face="bold"),
           axis.text.x=element_text(angle=45,hjust=1))
print(g)
```

As you may see tornados are the most hazardous phenomena while it's not the most common event type. Alnog with top ten events by damage we selected top 10 most often weather events. The inttersection of those two sets are in Table 1. 
```{r evtable, }
bycount<-events[head(order(events$occurs,decreasing = TRUE),10),]
topinjures <- dplyr::intersect(bycount,byharm)
kable(topinjures,col.names = c("Event type","Injured","Total events"),caption = "Table 1. The most harmful and common weather events.")
```

## Property damage
